.se title   = 'NetRexx User''s Guide'
.se stitle  = 'User''''s Guide'      ;.* '''' as re-embedded below
.se status  = 'Version'
.se version = '2.00'
.se date    = '31st August 2000'
.se vdate   = '31 Aug 2000'          ;.* must be date() format
.se stddate = 'Aug 2000'
.*
.se final   = '0'                    ;.* 1 for final print (no footings)
.se startpag= '0'                    ;.* 1 to include start page
.se imsupp  = '1'                    ;.* 1 for supplement
.se imindex = '1'                    ;.* 1 to include Index
.se copyear = '2000'                 ;.* copyright year
.*
.*
:gdoc    sec='&status. &version.'.
.**********************************************************************
.*
.*  "The NetRexx User's Guide"
.*
.*  (C) IBM UK Laboratories Ltd./Mike Cowlishaw 1996-2000
.*
.**********************************************************************
.*
.*  This imbeds SMFF directly, rather than using GML4PLUS (which
.*    breaks, nowadays)
.*
.**********************************************************************
.* Next should be null for final print
.se footstatus =''
.se footcopy =''
.if &final = 1 .go nofoots
.se footstatus ='&status. &version.'
.se footcopy ='Copyright (c) IBM Corporation 1996, &copyear..  All rights reserved.'
... nofoots
.* - - - - - - - - - - - - - - - - - - - - - - - -
.'im omnisym    ;.* Omnisym (for &Ea, etc.)
.'im omsymext   ;.* Omnisym extensions
.'im dsmfprof   ;.* Script Mathematical Formula Formatter
.'im nrlprof    ;.* SCRIPT profile for this document
.'im decprof    ;.* Modifications as used for Decimal documents
.* .'im nrlph   ;.* The Prentice-Hall style implementation
.* ----------------------------------------------------------------
:frontm.
:titlep.
:title stitle='&stitle.'.&title.
:date.&date.
:author.Mike Cowlishaw
:address.
:aline.mfc@uk.ibm.com
:aline.IBM UK Laboratories
:eaddress.
:etitlep.
.*----------- The Front Matter -----------------------------------------
.* Currently just TOC
:toc.
.*----------- The Body -------------------------------------------------
:body.
.dh 0 nodot     ;.* No dots in contents
.dh 1 nodot
.dh 2 nodot pa
.dh 3 nodot tc
.dh 4 nodot ntc
.*----- The true body follows from master -----
