\chapter{Installing and running on the Raspberry Pi}
\subsection{Running NetRexx in 10 minutes on the Raspberry Linux/ARM system}
\index{Raspberry Pi}
\index{installation,Raspberri Pi}
This install guide is different, in the sense that it describes the entire setup of the Raspberry Pi system, including NetRexx.

\subsubsection{Linux on ARM}
The Raspberry Pi is an inexpensive computer, containing an ARM architecture CPU on a board the size of a credit card, which sells for \$35. It boots from an SD card, the kind you have in your digital camera. In a few small steps you can be up and running with Java and NetRexx. 
\begin{itemize}
\item For speed, resist the urge to unpack the Raspberry, but do unpack an SD card of suitable size (and known brand)\footnote{Not all cards work; the large brands do.}, at least 2GB but 8 or 16 is advisable
\item Download the raspbian image from \url{http://www.raspberrypi.org/downloads}, or alternatively, the soft float wheezy image if you are mainly going to use this for java\footnote{For the time being, the soft float JVM from Oracle is going to be faster because the JIT compiler works; it does not on the hard float raspbian image (which overall is faster). This will be corrected soon, no doubt, by an Oracle hard float version that can JIT.}
\item Hook up an SD Card writer (the one in your digital camera probably also works) to the USB port of your computer
\item While taking good care not to overwrite your harddisk, use \emph{dd} or, on Windows, \emph{Win32DiskManager} to write the image to the SD card. This takes a minute. Good instructions are at \url{http://elinux.org/RPi_Easy_SD_Card_Setup}
\item Now unpack the Raspberry Pi, connect the hdmi to a tv or via an hdmi-monitor cable to a monitor, connect a keyboard (mouse can be attached later, if at all), and connect the mini-usb adapter to the power socket. I used a spare plug from an old phone. It boots and gives a lot of Unix messages. The first boot is not very quick. Connect an ethernet cable to your router.
\item You land in the raspi-config system. Resize the partitions to fill your SD card. Change the password for the pi user, set the default locale, and enable ssh. You can worry with the other options later.
\item Note the IP address that the system received from DHCP
\item Login from another system, for example using Putty (for Windows) or use ssh pi@your.ip.add.ress (these are the numbers of an IP4 address)
\item When logged on, issue \texttt{sudo su -}. You are root now; be careful.
\item Issue \texttt{aptitude install openjdk-6-jdk}. Enter through the prompts. Alternatively, download the ARM Java JDK from Oracle for use on the soft float wheezy debian image, and untar it.
\item Use scp or ftp (binary mode) to transmit the NetRexxC.jar to the system, or install the whole NetRexx package. There is an unzip command available
\item Set path and classpath as indicated earlier. And run NetRexx. It is possible to develop on the Raspberry, or just upload class files to it. It has a 700Mhz CPU, it is good to adapt your expectations of its performance to this fact; at least it will have the best performance per dollar for some time.
\end{itemize}