\input{../../../boilerplate/preamble}
\begin{document} 
\renewcommand{\isbn}{978-90-819090-0-6}
\setcounter{tocdepth}{1}
\title{\fontspec{Bodoni URW Light}NetR\textsc{exx}\protect\\\fontspec{Bodoni URW Light}Programming Guide}
\author{RexxLA}
\date{\null\hfill Version \splice{java org.netrexx.process.NrVersion} of \today}
\maketitle
\pagenumbering{Roman}
\pagestyle{fancy}
\frontmatter
\pagenumbering{Roman}
\input{../../../boilerplate/bookmeta}
\input{../../../boilerplate/hyphenation}
\tableofcontents
\newpage
\pagenumbering{arabic}
\frontmatter
\large
\input{../../../boilerplate/series}
%\input{../boilerplate/conventions}
\include {introduction}
\mainmatter
\chapter{Meet the \Rexx{} Family}
\section{Once upon a Virtual Machine}
On the 22nd of March 1979 Mike Cowlishaw of IBM had a
vision of an easier to use command processor for VM, and wrote down a
specification over the following days.
VM\textsuperscript{\texttrademark} (now called z/VM) is the original Virtual Machine operating system,  stemming from an
era in which time sharing was acknowledged to be the wave of the
future and when systems as CTSS (on the IBM 704) and TSS (on the IBM
360 Family of computers) were early timesharing systems, that offered
the user an illusion of having a large machine for their exclusive
use, but fell short of virtualising the entire hardware. The CP/CMS
system changed this; CP virtualised the hardware completely and CMS was the OS
running on CP. CMS knew a succession of command interpreters, called
EXEC, EXEC2 and \Rexx{}\textsuperscript{\texttrademark}  (originally REX
- until it was found out, by the IBM legal department, that a product of another vendor had a similar name) -
the EXEC roots are the explanation why some people refer to a \nr{}
program as an ``exec''. As a prime example of a \emph{backronym}, Rexx
stands for ``Restructured Extended Executor''. It
can be defended that \Rexx{} came to be as a reaction on EXEC2, but it
must be noted that both command interpreters shipped around the same
time. From 1988 on \Rexx{} was available on MVS/TSO and other systems,
like DOS, Amiga and various Unix systems. \Rexx{} was branded the
official SAA procedures language and was implemented on all IBM's
Operating Systems; most people got to know \Rexx{} on OS/2. In the late
eighties the Object-Oriented successor of Rexx, Object Rexx, was
designed by Simon Nash and his colleagues in the IBM Winchester
laboratory. \Rexx{} was thereafter known as Classic Rexx. Several open
source versions of Classic \Rexx{} were made over the years, of which
Regina is a good example.

\section{Once upon another Virtual Machine}
In 1995 Mike Cowlishaw ported Java\texttrademark to OS/2\textsuperscript{\texttrademark} and soon after started with
an experiment to run \Rexx{} on the JVM\textsuperscript{\texttrademark}. With \Rexx{} generally considered
the first of the general purpose scripting languages, \nr{}\textsuperscript{\texttrademark}  is the
first alternative language for the JVM. The 0.50 release, from April
1996, contained the \nr{} runtime classes and a translator written
in \Rexx{} but tokenized and turned into an OS/2 executable. The 1.00
release came available in January 1997 and contained a translator
bootstrapped to \nr{}. The \Rexx{} string type that can also handle
unlimited precision numerics is called \Rexx{} in Java and \nr{}.
Where Classic \Rexx{} was positioned as a system \emph{glue} language and
application macro language, \nr{} is seen as the one language that
does it all, delivering system level programs or large applications.

Release 2.00 became available in August 2000 and was a major upgrade,
in which interpreted execution was added. Until that release, \nr{}
only knew \emph{ahead of time} compilation (AOT).

Mike Cowlishaw took early retirement from IBM in March 2010. IBM
announced the transfer of \nr{{} source code to the \Rexx{} Language
  Association (RexxLA) on June 8, 2011, 14 years after the v1.0
  release, and on the same day, it released the \nr{} source code to RexxLA
under the ICU open source license. RexxLA shortly after released this
as \nr{} 3.00 and has followed with updates.
\section{Features of \nr}
\begin{description}
\item[Ease of use]
The \nr{} language is easy to read and write because many instructions
are meaningful English words. Unlike some lower level programming
languages that use abbreviations, \nr{} instructions are common words, such as \textbf{say}, \textbf{ask}, \textbf{if...then...else}, \textbf{do...end}, and \textbf{exit}.
\item[Free format] There are few rules about \nr{} format. You need not start an instruction in a particular column, you can also skip spaces in a line or skip entire lines, you can have an instruction span many lines or have multiple instructions on one line, variables do not need to be pre-defined, and you can type instructions in upper, lower, or mixed case.
\item[Convenient built-in functions] \nr{} supplies built-in functions
  that perform various processing, searching, and comparison
  operations for both text and numbers. Other built-in functions
  provide formatting capabilities and arithmetic calculations.
\item[Easy to debug]
When a \nr{} exec contains an error, messages with meaningful explanations are displayed on the screen. In addition, the \textbf{trace} instruction provides a powerful debugging tool.
\item[Interpreted]
The \nr{} language is an interpreted language. When a \nr{} exec
runs, the language processor directly interprets each language
statement, or translates the program in JVM bytecode.
\item[Extensive parsing capabilities]
\nr{} includes extensive parsing capabilities for character
manipulation. This parsing capability allows you to set up a pattern
to separate characters, numbers, and mixed input.
\item[Seamless use of JVM Class Libraries]
\nr{} can use any class, and class library for the JVM (written in
Java or other JVM languages) in a seamless manner, that is, without
the need for extra declarations or definitions in the source code.
\end{description}


\chapter{Learning to program}
\section{Console Based Programs}
One way that a computer can communicate with a user is to ask
questions and then compute results based on the answers typed in. In
other words, the user has a conversation with the computer. You can
easily write a list of  \nr{} instructions that will conduct a conversation. We call such a list of instructions a program.
The following listing shows a sample \nr{} program. The sample program asks the user to give his name, and then responds to him by name. For instance, if the user types in the name Joe, the reply Hello Joe is displayed. Or else, if the user does not type anything in, the reply Hello stranger is displayed.
First, we shall discuss how it works; then you can try it out for
yourself.
\begin{lstlisting}[label=hello,caption=Hello Stranger]
/* A conversation */
say "Hello! What's your name?"
who=ask
if who = '' then say "Hello stranger"
else say "Hello" who
\end{lstlisting}
Briefly, the various pieces of the sample program are:
\begin{description}
\item[\texttt{/* ... */}] A comment explaining what the program is
  about. Where \Rexx{} programs on several platforms must start with a comment, this is not
  a hard requirement for \nr{} anymore. Still, it is a good idea to start
  every program with a comment that explains what it does.
\item [\texttt{say}] An instruction to display Hello! What' s your name? on the screen.
\item [\texttt{ask}] An instruction to read the response entered from the keyboard and put it into the computer's memory.
\item [\texttt{who}] The name given to the place in memory where the user's response is put.
\item [\texttt{if}] An instruction that asks a question. 
\item [\texttt{who = ''}] A test to determine if who is empty.
\item [\texttt{then}] A direction to execute the instruction that follows, if the tested condition is true.
\item [\texttt{say}] An instruction to display Hello stranger on the screen.
\item [\texttt{else}] An alternative direction to execute the
  instruction that follows, if the tested condition is not true. Note
  that in \nr{}, else needs to be on a separate line.
\item [\texttt{say}] An instruction to display Hello, followed by whatever is in who on the screen.
\end{description}
The text of your program should be stored on a disk that you have
access to with the help of an \emph{editor} program. On Windows,
notepad or (notepad++), jEdit, X2 or SlickEdit are suitable
candidates. On Unix based systems, including macOS, vim or emacs are
plausible editors. If you are on z/VM or z/OS, XEDIT or ISPF/PDF are a
given. More about editing \nr{} code in chapter \ref{editors},
\emph{Editor Support}, on page \pageref{editors}. 

When the text of the program is stored in a file, let's say we called
it \texttt{hello.nrx}, and you installed \nr{} as indicated in the
\emph{\nr{} QuickStart Guide}, we can run it with
\begin{verbatim}
    nrc -exec hello
\end{verbatim}
and this will yield the result:
\begin{alltt}
\nr{} portable processor, version \nr{} after3.01, build 1-20120406-1326
Copyright (c) RexxLA, 2011.  All rights reserved.
Parts Copyright (c) IBM Corporation, 1995,2008.
Program hello.nrx
===== Exec: hello =====
Hello! What’s your name?
\end{alltt}
If you do not want to see the version and copyright message every
time, which would be understandable, then start the program with:
\begin{verbatim}
nrc -exec -nologo hello
\end{verbatim}
This is what happened when Fred tried it.
\begin{verbatim}
Program hello.nrx
===== Exec: hello =====
Hello! What’s your name?
Fred
Hello Fred
\end{verbatim}
The \textbf{ask} instruction paused, waiting for a reply. Fred typed
Fred on the command line and, when he pressed the ENTER key, the
\textbf{ask} instruction put the word Fred into the place in the
computer's memory called ``who''. The \textbf{if} instruction asked,
is ``who'' equal to nothing:
\begin{verbatim}
who = '' 
\end{verbatim}
meaning, is the value of ``who''  (in this case, Fred) equal to
nothing:
\begin{verbatim}
"Fred = ''
\end{verbatim}
This was not true; so, the instruction after \texttt{then} was not executed; but the instruction
after \texttt{else}, was.

But when Mike tried it, this happened:
\begin{verbatim}
Program hello.nrx
===== Exec: hello =====
Hello! What’s your name?

Hello stranger
Processing of 'hello.nrx' complete
\end{verbatim}
Mike did not understand that he had to type in his name. Perhaps the
program should have made it clearer to him. Anyhow, he just pressed
ENTER. The \textbf{ask} instruction put '' (nothing) into the place in the
computer's memory called ``who''. The \textbf{if} instruction asked, is:
\begin{verbatim}
who = ''
\end{verbatim}
meaning, is the value of ``who'' equal to nothing:
\begin{verbatim}
'' = ''
\end{verbatim}
 
In this case, it was true. So, the instruction after \textbf{then} was
executed; but the instruction after \textbf{else} was not.

\section{Comments in programs}
When you write a program, remember that you will almost certainly want to read it over later (before improving it, for example). Other readers of your program also need to know what the program is for, what kind of input it can handle, what kind of output it produces, and so on. You may also want to write remarks about individual instructions themselves. All these things, words that are to be read by humans but are not to be interpreted, are called comments.
To indicate which things are comments, use:
\begin{verbatim}
/* to mark the start of a comment 
*/ to mark the end of a comment.
\end{verbatim}
The \texttt{/*} causes the translator to stop compiling and interpreting;
this starts again only after a \texttt{*/} is found, which may
be a few words or several lines later. For example,
\begin{verbatim}
/* This is a comment. */
say text /* This is on the same line as the instruction */
/* Comments may occupy more
than one line. */

\end{verbatim}
\nr{} also has line mode comments - those turn a line at a time into
a comment. They are composed of two dashes (hyphens, in listings sometimes
fused to a typographical \emph{em dash} - remember that in reality
they are two \emph{n dashes}.
\begin{verbatim}
-- this is a line comment
\end{verbatim}
\section{Strings}
When the translator sees a quote (either " or ') it stops
interpreting or compiling and just goes along looking for the matching quote. The
string of characters inside the quotes is used just as it is. Examples
of strings are:
\begin{verbatim}
'Hello'
"Final result: "
\end{verbatim}
If you want to use a quotation mark within a string you should use
quotation marks of the other kind to delimit the whole string.
\begin{verbatim}
"Don't panic"
'He said, "Bother"'
\end{verbatim}
There is another way. Within a string, a pair of quotes (of the same
kind as was used to delimit the string) is interpreted as one of that
kind.
\begin{verbatim}
'Don''t panic' (same as "Don't panic" )
 "He said, ""Bother""" (same as 'He said, "Bother"')
\end{verbatim}
\section{Clauses}
Your \nr{} program consists of a number of \emph{clauses}. A clause
can be:
\begin{enumerate}
 \item A \emph{keyword instruction} that tells the interpreter to do something; for
   example,
\begin{verbatim}
say  "the word"
\end{verbatim}
In this case, the interpreter will display the word on the user's
screen. 
\item An \emph{assignment}; for example,
\begin{verbatim}
Message = 'Take care!'
\end{verbatim}
\item A \emph{null} clause, such as a completely blank line, or
\begin{verbatim}
    ;
\end{verbatim}
\item A \emph{method call instruction} which invokes a \emph{method}
    from a \emph{class}
\begin{verbatim}
'hiawatha'.left(2)
\end{verbatim}
\end{enumerate}
\section{When does a Clause End?}
It is sometimes useful to be able to write more than one clause on a
line, or to extend a clause over many lines. The rules are:
\begin{itemize}
\item Usually, each clause occupies one line.
\item If you want to put more than one clause on a line you must use a semicolon (;) to separate the clauses.
\item If you want a clause to span more than one line you must put a
  dash (hyphen) at the end of the line to indicate that the clause
  continues on the next line. If a line does not end in a dash, a
  semicolon is implied.
\end{itemize}
What will you see on the screen when this exec is run?
\begin{lstlisting}[label=rah,caption=RAH Exec]
/* Example: there are six clauses in this program */ say "Everybody cheer!"
say "2"; say "4" ; say "6" ; say "8" ; say "Who do we" -
"appreciate?"
\end{lstlisting}
\section{Long Lines}
Ever since the days of the punch card images are over, the lines in program
sources have become longer and longer, and with \nr{} being a free
format language, there is no real technical reason to limit line
length. Still, for readability and for ease access to words within a
line, it is often indicated to keep lines relatively short and
tidy. For this reason, the \emph{continuation character} '-' can be
used. This also makes it possible to split long literal strings over
lines.
\begin{lstlisting}[label=longline,caption=Long lines]
say 'good' - 
'night'
\end{lstlisting}
This example will concatenate 'good' and 'night' with a space
inbetween. When you want to avoid that, use the '||' concatenation
operator.
\begin{lstlisting}[label=longlineconcat,caption=Long lines with string
  concatenation without space]
say 'good' - 
||'night'
\end{lstlisting}

\section{Loops}
We can go on and write clause after clause in a program source files,
but some repetitive actions in which only a small change occurs, are
better handled by the \textbf{loop} statement.
% .reminding
% of an anecdote that Andy Hertzfield tells\footnote{\url{http://www.folklore.org}:
% Bob's background looked to be a lot stronger in hardware than software, so we were somewhat skeptical about his software expertise, but he claimed to be equally adept at both. His latest project was a rebellious, skunk-works type effort to make a low cost version of the Star called "Cub" that used an ordinary Intel microprocessor (the 8086), which was heresy to the PARC orthodoxy, who felt that you needed custom, bit-slice processors to get sufficient performance for a Star-type machine. Bob had written much of the software for Cub himself. 

% "I've got lots of software experience", he declared, "in fact I've personally written over 350,000 lines of code." 

% I thought that was pretty impressive, although I wondered how it was calculated. I couldn't begin to honestly estimate how much code I have written, since there are too many different ways to construe things. 

% That evening, I went out to dinner with my friend Rich Williams, who started at Apple around the same time that I did. Rich had a great sense of humor. I told him about the interview that I did in the afternoon, and how Bob Belleville claimed to have written over 350,000 lines of code. 

% "Well, I bet he did", said Rich, "but then he discovered loops!"}
Imagine an assignment to neatly print out a table of exchange rates for
dollars and euros for reference in a shop. We could of course make the
following program:
\begin{lstlisting}[label=withoutloop,caption=Without a loop]
say  1 'euro equals'  1  * 1.19 'dollars'
say  2 'euro equals'  2  * 1.19 'dollars'
say  3 'euro equals'  3  * 1.19 'dollars'
say  4 'euro equals'  4  * 1.19 'dollars'
say  5 'euro equals'  5  * 1.19 'dollars'
say  6 'euro equals'  6  * 1.19 'dollars'
say  7 'euro equals'  7  * 1.19 'dollars'
say  8 'euro equals'  8  * 1.19 'dollars'
say  9 'euro equals'  9  * 1.19 'dollars'
say 10 'euro equals' 10  * 1.19 'dollars'
\end{lstlisting}
This is valid, but imagine the alarming thought that the list is deemed
a success and you are tasked with making a new one, but now with
values up to 100. That will be a lot of typing.

The way to do this is using the \textbf{loop}\footnote{Note that
  Classic \Rexx{} uses \textbf{do} for this purpose. In recent Open
  Object \Rexx{} versions
  \textbf{loop} can also be used.} statement.
\begin{lstlisting}[label=withoop,caption=With a loop]
loop i=1 to 100
  say i 'euro equals' i * 1.19 'dollars'
end
\end{lstlisting}
Now the \emph{loop index variable} \texttt{i} varies from 1 to 100,
and the statements between \texttt{loop} and \texttt{end} are
repeated, giving the same list, but now from 1 to 100 dollars.

We can do more with the \textbf{loop} statement, it is extremely
flexible. The following diagram is a (simplified, because here we left
out the \emph{catch} and \emph{finally} options) rundown of the ways
we can loop in a program.
%% \input{../../../railroad/loop}

A few examples of what we can do with this: 
\begin{itemize}
\item Looping forever - better put, without deciding beforehand how
  many times
\begin{lstlisting}[label=loopforever,caption=Loop Forever]
loop forever
  say 'another bonbon?'
  x = ask
  if x = 'enough already' then leave
end
\end{lstlisting}
The \texttt{leave} statement breaks the program out of the loop. This
seems futile, but in the chapter about I/O we will see how useful this
is when reading files, of which we generally do not know in advance
how many lines we will read in the loop.

\item Looping for a fixed number of times without needing a loop index
  variable
\begin{lstlisting}[label=loopfixed,caption=Loop for a fixed number of
  times without loop index variable]
loop for 10
   in.read() /* skip 10 lines from the input file */
end
\end{lstlisting}
\item Looping back into the value of the loop index variable
\begin{lstlisting}[label=loopforever,caption=Loop Forever]
loop i = 100 to 90 by -2
  say  i
end
\end{lstlisting}
This yields the following output:
\begin{verbatim}
===== Exec: test =====
100
98
96
94
92
90
Processing of 'test.nrx' complete
\end{verbatim}
\end{itemize}

\section{Special Variables}
We have seen that a \emph{variable} is a place where some data, be it
character date or numerical data, can be held. There are some special
variables, as shown in the following program.
\lstinputlisting[label=specialvars,caption=\nr{} Special Variables]{../../../../examples/rosettacode/RCSpecialVariables.nrx}
\begin{description}
\item[\texttt{this}] The special variables \textbf{this} and \textbf{super} refer to the
current instance of the class and its superclass - what this means
will be explained in detail in the chapter \textbf{Classes} on page
\pageref{classes}, as is the case with the \textbf{class} variable.

\item[\texttt{digits}]The special variable \textbf{digits} shows the current setting for the
number of decimal digits - the current setting of \textbf{numeric
  digits}. The related variable \textbf{form} returns the current
setting of \textbf{numeric form} which is either \texttt{scientific}
or \texttt{engineering}.

\item[\texttt{null}]The special variable \textbf{null} denotes the \emph{empty
  reference}. It is there when a variable has no value.

\item[\texttt{source}]The \textbf{source} and \textbf{sourceline} variables are a good way to
show the sourcefile and sourceline of a program, for example in an
error message.

\item[\texttt{trace}]The \textbf{trace} variable returns the current trace setting, which
can be one of the words \texttt{off var methods all results}. 

\item[\texttt{version}]The \textbf{version} variable returns the version of the \nr{}
translator that was in use at the time the clause we processed; in
case of interpreted execution(see chapter \ref{interpreted} on \pageref{interpreted}, it returns
the level of the current translator in use. 
\end{description}
The result of executing this exec is as follows:
\begin{alltt}
===== Exec: RCSpecialVariables =====
<super>RCSpecialVariables@4e99353f</super>
<this>RCSpecialVariables@4e99353f</this>
<class>class RCSpecialVariables</class>
<digits>9</digits>
<form>scientific</form>
<[1, 2, 3].length>
3
</[1, 2, 3].length>
<null>

</null>
<source>Java method RCSpecialVariables.nrx</source>
<sourceline>21</sourceline>
<trace>off</trace>
<version>\nr{} 3.02 27 Oct 2011</version>
Type an answer:
hello fifi
<ask>hello fifi</ask>
\end{alltt}
It might be useful to note here that these special variables are not
fixed in the sense of that they are not \emph{Reserved Variables}. \nr{}
does not have reserved variables and any of these special variables
can be used as an ordinary variable. However, when it is used as an
ordinary variable, there is no way to retrieve the special behavior.

\include{nrscripting}
\chapter{\nr{} as an Interpreted Language}\label{interpreted}
In the JVM environment, compilation and interpretation are concepts
that are not as straightforward as in other environments; JVM code is
interpreted on several levels. When we are referring to
\emph{interpreted} \nr{} code, we indicate that there is no
intermediate Java compilation step involved. A JVM .class file is
always interpreted by the JVM runtime; the \nr{} translator is able to
execute programs without generating either .java or .class files.

This enables a very quick edit-debug-run cycle, especially when
combined with the command line feature that keeps the translator
classes resident (the -prompt option), or one of the IDE plugins for
\nr{}.

For \nr{} to deliver this functionality, the translator has been
designed to have an analogous interpret facility for every code
generation part.\footnote{This is the right order in which to explain this
  feature, because historically, the compiler was first (1996) and the
  interpretation facility was added later (in 2000) -(but not without an
  extensive redesign of the compiler).}

\include{RxModel}

\include{nrucomp}
% \chapter{\nr{} as a Compiled Language}
% \input{nrucomp}
% The translator may be called from a \nr{} or Java program directly, by
% invoking the \emph{main} method in the \emph{org.netrexx.process.NetRexxC}
% class described as follows:

% \begin{lstlisting}[label=ivmain,caption=Invoking NetRexxC.main]
% method main(arg=Rexx, log=PrintWriter null) static returns int
% \end{lstlisting}

% The \emph{Rexx} string passed to the method can be any combination of
% program names and options (except \emph{-run}), as described above.
% Program names may optionally be enclosed in double-quote characters (and
% must be if the name includes any blanks in its specification).

% A sample \nr{} program that invokes the \nr{} compiler to compile a
% program called \emph{test} is:
% \begin{lstlisting}[label=compiletest,caption=Compiletest]
% /* compiletest.nrx */
% s='test -keep -verbose4 -utf8'
% say org.netrexx.process.NetRexxC.main(s)
% \end{lstlisting}

% Alternatively, the compiler may be called using the method:
% \begin{lstlisting}[label=array,caption=Calling with Array argument]
% method main2(arg=String[], log=PrintWriter null) static returns int
% \end{lstlisting}
% in which case each element of the \emph{arg} array must contain
% either a name or an option (except \emph{-run}, as before).  In this
% case, names must \emph{not} be enclosed in double-quote characters, and
% may contain blanks.

% \index{completion codes, from translator}
% \index{return codes, from translator}
% \index{PrintWriter stream for capturing translator output}
% \index{capturing translator output}
% For both methods, the returned \emph{int} value will be one of the
% return values described above, and the second argument to the method is
% an optional \emph{PrintWriter} stream.  If the \emph{PrintWriter} stream
% is provided, translator messages will be written to that stream (in
% addition to displaying them on the console, unless \emph{-noconsole} is
% specified).
% It is the responsibility of the caller to create the stream (autoflush
% is recommended) and to close it after calling the compiler.
% The \emph{-savelog} compiler option is ignored if a \emph{PrintWriter}
% is provided (the \emph{-savelog} option normally creates
% a \emph{PrintWriter} for the file \emph{NetRexxC.log}).

% \textbf{Note:} NetRexxC is thread-safe (the only static properties are constants), but
% it is not known whether \emph{javac} is thread-safe.  Hence the
% invocation of multiple instances of NetRexxC on different threads should
% probably specify \emph{-nocompile}, for safety.

% \section{Compiling from memory strings}
% Programs may also be compiled from memory strings by passing an array
% of strings containing programs to the translator using these methods:

% \begin{lstlisting}[label=frommemory,caption=From Memory]
% method main(arg=Rexx, programarray=String[], log=PrintWriter null) static returns int
% method main2(arg=String[], programarray=String[], log=PrintWriter null) static returns int
% \end{lstlisting}

% Any programs passed as strings must be named in the arg parameter before any programs contained in files are named.
% For convenience when compiling a single program, the program can be
% passed directly to the compiler as a String with this method:

% \begin{lstlisting}[label=string,caption=With String argument]
% method main(arg=Rexx, programstring=String, logfile=PrintWriter null) constant returns int
% \end{lstlisting}

% Here is an example of compiling a \nr{} program from a string in
% memory:

% \begin{lstlisting}[label=memexample,caption=Example of compiling from String]
% import org.netrexx.process.NetRexxC
% program = "say 'hello there via NetRexxC'"
% NetRexxC.main("myprogram",program)
% \end{lstlisting}

% \marginnote{\color{gray}3.01}}Programs may also be interpreted directly from memory strings, as
% shown in \ref{interpretstrings} on page \pageref{interpretstrings}.



\include{nruapi}

\chapter{Calling non-JVM programs}
Non-JVM programs can be called using the \keyword{Address} instruction.
For optimal flexibility in the handling of output, this sections
describes how to use the native Java facilities for this.
It is easy to call non-JVM programs
from a \nr{} program - not as easy as calling a JVM class of course, but if
the following recipe is observed, it will show not to be a major
problem. The following example is reusable for many cases.
\lstinputlisting[label=nonjava,caption=Calling Non-JVM
Programs]{../../../../examples/NrxRedBk/script/NonJava.nrx} 
Just firing off a program is no big deal, but this example (in script
style) shows how easy it is to access the in- and output handles for
the environment that executes the program, which enables you to
capture the output the non-jvm program produces and do useful things
with it.\footnote{This is akin to what one would do with \emph{queue}
  on z/VM CMS and \emph{outtrap} on z/OS TSO in Classic Rexx.}
Line 17 starts the external command using the JVM \texttt{Runtime}
class in a process called \texttt{child}. In line 20 we create a
\texttt{BufferedReader} from the \texttt{child} processes'
output. This is called an InputStream but it might as well have been
called an OutputStream- everything regarding I/O is relative - but
fortunately the designers of the JVM took care of deciding this for you.
In lines 25-35 we loop through the results and show the files stored
in the zipfile. Note that we \textbf{do} (line 14) have to \textbf{catch} (line
42) the \emph{IOException} that ensues if the runtime cannot find the
\texttt{unzip} program, maybe because it is not on the path or was not
delivered with your operating system. 


Starting from JVM 1.5 releases, there is a new way to accomplish the
same goal, in a cleaner manner and with the added bonus of being able
to redirect streams, and use environment variables. In this regard,
the environment variable has made an important comeback from having
its calls deprecated, to easy to use support in the
\emph{ProcessBuilder} class. 
\lstinputlisting[label=nonjava,caption=Use of ProcessBuilder]{../../../../src/org/netrexx/address/OSProcess.nrx} 
In the above sample, we are using two different ways to obtain the
output from a process started by the JVM from our own program. The
method \emph{outtrap} waits until the invoked process is finished and
returns all output lines in an \texttt{ArrayList}. Its name is not
entirely coincidental with the similar TSO outtrap function. 

Sometimes we cannot wait until the child process is finished, for
example when it is a long running process and we need to capture the
output on a line-by-line basis to see what is happening - in case of
the example, this was done to capture the output as part of a
testsuite for a multithreaded file transfer application, which has a
server resident process that is not supposed to end, because one of
its tasks is to poll a directory for incoming files with a specific
pattern in the file names. This is implemented using an Event based
pattern (as explained in \ref{events}  on page \pageref{events}. 
\lstinputlisting[label=nonjava,caption=Output Line
Event]{../../../../src/org/netrexx/address/OutputLineEvent.nrx} 

\lstinputlisting[label=outputeventlistener,caption=Output Event Listener]{../../../../src/org/netrexx/address/OutputEventListener.nrx} 

The call would look something like this:
\begin{lstlisting}[label=callosprocessexample,caption=Example of calling the OSProcess class - registering an eventhandler]
    os = OSProcess()
    os.addOutputEventListener(this)
    os.exec(command)
\end{lstlisting}

The class must \texttt{extend OutputEvenListener}, and implement this
method:

\begin{lstlisting}[label=registerhandlerexample,caption=Example of implementing the listener method]
  method outputReceived(ob=OutputLineEvent)
    this.counter = this.counter+1
    say this.counter ob.getPid() ob.getLine()
\end{lstlisting}

\chapter{Using \nr{} classes from Java}
If you are a Java programmer, using a \nr{} class from Java is just as
easy as using a Java class from \nr{}. \nr{} compiles to Java classes that can be used by Java programs.
You should import the netrexx.lang package to be able to use the short
class name for the \Rexx{} (\nr{} string and numerics) class.

A \nr{} method without a returns keyword can return nothing, which is the void type in Java, or a \Rexx{} string. 
\nr{} is case independent\footnote{With the default of \texttt{options
  nostrictcase} in effect.}; Java is case dependent. \nr{} generates
the Java code with the case used in the class and method
instructions. For example, if you named your class Spider in the \nr{}
source file, the resulting Java class file is Spider.class.
The public class name in your source program must match the \nr{}
source file name. For example, if your source file is \texttt{SPIDER.NRX}, and
your class is \texttt{Spider}, \nr{} generates a warning and changes the
class name to \texttt{SPIDER} to match the file name. A Java program using the
class name Spider would not find the generated class, because its name
is \texttt{SPIDER.class} - if the compile succeeded, which is not guaranteed in
case of casing mismatches.
If you have problems, compile your \nr{} program with the \textbf{options
-keepasjava -format}. You then can look at the generated java file for the correct spelling style and method parameters.

\chapter{Classes}\label{classes}
Somewhere in the nineties Object Orientation became one of the
mainstream ways to organize computer programs, and support for this
was added to programming languages. C became C++ with a preprocessor
that generates C\footnote{Cfront} that is not entirely unlike the
  \nr{} translator produces Java. Java in itself is syntax-wise a
  cleaned up version of C++, but in essence an entirely different
  language. Its inventor and architect, James Gosling, has stated on
  various occasions that he was planning a fully different syntax for
  what finally became Java - but that Sun management more or less
  forced him to use a C++ derived syntax, because C++ compilers was
  what SUN did well at the time. With Brendan Eich having to base
  JavaScript qua naming and syntax on Java, the circle that brought
  the world terse, curly braces based notations, is complete.

For an audience of \Rexx{} programmers, the usual OO presentation goes
into the advantages of the paradigm. Today, that is not really
necessary, and OO is a given; it slightly deviates from earlier
notation as result of trying to put data and procedure into
\emph{Objects}, but it is no great deal, and this \nr{} Programmer's
Guide does not need a special section on the benefits of the OO
paradigm. It is assumed that with a few examples everyone should be
able to \emph{get} it; some old programmers might resist but there is
really no use in fighting the mainstream. Consequently, this section
discusses the way to do this in \nr{}; the way \nr{} does it is for a very
large part formed by the way the JVM dictates it, adapted to Rexx
notational style and conventions.

Where traditional \Rexx{} would say:
\begin{alltt}
l=left(ourstring,1)
\end{alltt}
the OO-versions of \Rexx{} would say:
\begin{alltt}
l=ourstring.left(1)
\end{alltt}
As often the case, the hard part is in the notational ommission that
OO has as its characteristic: the instance pointer is no part of the
function call and has moved to the left (in what now is called a
\emph{method}. The weight has shifted from the operation to the object
it is called on.
\section{Classes}
Classes represent a blueprint, 'cookie cutter' approach in creating
objects that do useful things. A class is defined in a file by the
same name (exceptions here for dependent classes). So a class called
Cookie is defined in a file called Cookie.nrx. Its \emph{real}, which
means its most specific name, including its package specification, is
not given by the file name but by the combination of the class=file +
the name given on the \texttt{package} statement. This enables one to
put classes in different packages without having to change the file
names.

\section{Dependent Classes}
Dependent classes are the \nr{} way to implement Java inner
classes. There is no in-line definition possible, and dependent
classes need their own class definition, but can be defined in the
same source file as the classes they depend on. The notational
advantage of 'nested' class definition, like customary in (for
example) Java Swing programs is absent. What is present, is the way
dependent classes can seamlessly access properties of their parent classes.
\section{Properties}
The properties statement enables us to define variables that are
global to the class definition, and as such can be used by all methods
of the class.
\begin{shaded}\noindent
A properties statement needs at least one \emph{visibility} or
\emph{modifier} keyword. When this is left out, a variable called
``properties'' is defined, which is not an error, but (most of the
times) not what was intended.
\end{shaded}\indent
Because the properties of a class can be externally visible (depending
on \emph{visibility} they need to have a type. When the type is
omitted in the definition, they are of type \keyword{Rexx}. So-called
\emph{indirect properties}, defined with the \keyword{properties
  indirect} modifier, give rise to automated generation of
\emph{getter} and \emph{setter} methods for use in Java Beans.
% \section{Methods}
% \section{Inheritance}
% \section{Overriding Methods}
% \section{Overriding Properties}
\chapter{Using Packages}
Any non-toy, non-trivial program needs to be in a package. Only
examples in programming books (present company included) have programs
without package statements. The reason
for this is that there is a fairly large chance that you will give
something a name that is already used by someone else for something
else. Things are not their names\footnote{Willard Van Orman Quine, Word
  and Object, MIT Press, 1960, ISBN 0-262-67001-1}, and the same names
are given to wildly dissimilar things. The \emph{package} construct is the JVM's approach to
introducing \emph{namespaces} into the total set of programs that
programmers make. Different people will probable write some method that is
called \texttt{listDifferences} sometime. With all my software in a
package called \texttt{com.frob.nitz} and yours in a package
called \texttt{com.frob.otzim}, there is no danger of our programs
calling the wrong class and listing the wrong differences.

It is imperative to understand this chapter before continuing - it is
a mechanical nuts-and-bolts issue but an essential one at that.

\section{The package statement}
The final words about the \nr{} \textbf{package} statement is in the
\nr{} Language Reference, but the final statement about the package
\emph{mechanism} is in the JVM documentation.
\section{Translator performance consequences}
Because the \nr{} translator has to scan all packages that it can
see (meaning a recursive scan of the directories below its own level
in the directory tree, and on its classpath, it is often advisable
(and certainly if . (a dot, representing the current directory) is part of the classpath)
to do development in a subdirectory, instead of, for example, the top
level home directory. If a large number of packages and classes are
visible to the translator, compile times will be negatively impacted. 

\section{Some \nr{} package history}
All IBM versions of \nr{} had the translator in a package called
\begin{verbatim}
COM.ibm.netrexx.process 
\end{verbatim}
The official, SUN ordained
convention for package names was, to prepend the reversed domain name
of the vendor to the package name, while uppercasing the top level
domain. \nr{}, being one of the first programs to make use of
packages, followed this convention, that was quickly dropped by SUN
afterwards, probably because someone experienced what trouble it could
cause with version management software that adapted to
case-\emph{sensitive} and case-\emph{insensitive} file systems. For
\nr{}, which had started out keenly observing the rules, this
insight came late. With the first RexxLA release of \nr{} in 2011, the package
name was changed to \texttt{org.netrexx}, while the runtime package
name was kept as \texttt{netrexx.lang}, also because some major other
languages follow this convention.

\include{jpms}

\chapter{Programming Patterns}
Much has been made of patterns as aggregations of higher level
embodiments of programming solutions.  It has been
observed\footnote{This observation from a Java patterns book.} that of a
number of the C++ oriented patterns in Design Patterns\footnote{Gamma,
Helm, Johnson, Vlissides, Design
Patterns: Elements of Reusable Object-Oriented Software,
Addison-Wesley Professional; 1994}, some owe their existence
to complications in the C++ language and are not readily reproducible
in a Java Patterns or Ruby Patterns book. The same goes for NetRexx-
in this chapter we would like to present a number of Java patterns
usable in NetRexx, and a number of patterns that are unique to
NetRexx.

\section{Singleton}\label{singleton}
Sometimes we only want one instance of a class, and we want every user
of the class to refer to that same instance. In this case, we need to
adapt the class construction mechanism to make sure this
happens. There are different ways to implement this, one way is shown
below.
\begin{lstlisting}[label=singleton1,caption=Singleton]
class TheGatherer

  properties static
  instance     = TheGatherer

  method getInstance() returns TheGatherer static protect
    if TheGatherer.instance <> null then
      do
	return TheGatherer.instance
      end
    else
      do
	TheGatherer.instance = TheGatherer()
	return TheGatherer.instance
      end
    
  /**
   * private constructor enforces singleton
   */
  method TheGatherer() private signals ClassNotFoundException
\end{lstlisting}
The way that has been chosen here is to make the constructor private,
so no other class can use it. We need an alternative method to make the first
and only instance of this class, and this is the \code{getInstance()}
method. This checks if a static property \code{instance} is null, in
which case the private constructor is run and its return value put in
\code{instance}. Every subsequent call to \code{getInstance()} sees
the value of the static variable instance being not null, and returns
that value, which now refers to the single instance. There are several
ways to enhance this method, but this is a simple way and it fits the
bill. For added security, override the methods for class
serialization.

There is a common naming pattern for Singletons, which is the prepend
the name of the class with \emph{The}, as in the above example.
\section{Observable and Events}\label{events}
The observer pattern can also be referred to as \emph{Callback}, and
the Java Event class delivers support for it. It is very usable if
some result needs to be available for a set of callers, where the set
is 0 to many. It works as follows: (see a simple implementation in
section \ref{outputeventlistener} on page
\pageref{outputeventlistener})
An object, maintains a list of its dependents, called observers, and
notifies them automatically of any state changes, usually by calling
one of their methods. It is mainly used to implement distributed event
handling systems. The Observer pattern is also a key part in the
familiar Model View Controller (MVC) architectural pattern. In the
JVM, this object needs to implement the methods of the Listener
interface; this interface specifies the addListener and RemoveListener
methods; it keeps a collection in which references to the added
listener objects are maintained. The listening is done to subclassed
Java Event classes. The event specifies the method to be called when
'firing off' and event. This means that this method is called on every
listener.

One of the larger benefits: it decouples the observer from the
subject. The subject doesn't need to know anything special about its
observers. Instead, the subject simply allows observers to
subscribe. When the subject generates an event, it simply passes it to
each of its observers. Another benefit is that event consuming classes
don't have to wait until a process is finished, and can consume events
as they come in. The OSProcess class on page
\pageref{outputeventlistener}) uses an event approach to consume
output lines from a subprocess - in the version that puts the output
in an ArrayList needs to wait for the subprocess to end, but the event
driven version can monitor a long running process and analyze output
lines whenever they are received.

\section{Recursive Parse}
This is a pattern somewhat unique to Rexx, by virtue of \Rexx{} having the Parse
statement. It also works in NetRexx.
\begin{lstlisting}
/* process a string word by word */

testString = "Foo Bar Baz Frob Frobnitz Frobbotzim"

loop until testString = ""

-- copy the first word of testString into curName
-- remove it from testString

  parse testString curName testString

  say "Current word is " curName
  say "Remaining words in testString are " testString
  end -- loop until testString
\end{lstlisting}
This enables one to 'peel off' one word of a sentence at at time.


\section{More Observer/Observable}

Java has special support for the Observer/Observable pattern in the
form of the \emph{Observer} class and the \emph{Observable} interface.
In the following snippet one can see the Observer class in working.

The class is the same \emph{singleton} as shown above, and starts several threads which need
to be observed.
\begin{lstlisting}[label=obs,caption=Observer]
class TheGatherer implements Observer 
\end{lstlisting}

The Observable threads need to implement the \emph{Observable}
interface, and to be able to be started as a thread, \emph{Runnable}.
This is how we start it; its definition follows.

\begin{lstlisting}[label=obs2,caption=Observable: starting Observable Threads]
  method TheGatherer() private signals ClassNotFoundException
    logger_.info( "TheGatherer: start")
    t1 = TransactionStatusMonitor(Rexx 10000)
    t1.addObserver(this)
    Thread(t1).start()
    logger_.info( "TheGatherer: started thread TransactionStatusMonitor")
\end{lstlisting}

We instantiate the Observable class  as \code{t1}, and add the instance of
our Observer class, TheGatherer, to it as Observer. After we have done
this, we start it by instantiating a Thread object with it and calling
the start method, which, by virtue of it implementing the
\emph{Runnable} interface, starts its \textbf{run} method.

Note that the TransactionStatusMonitor extends a class called
\code{Monitor}, which in turn, implements the \code{Observable}
interface. The reason for this is, we run several monitoring threads,
and they all behave in the same way.
\begin{lstlisting}[label=obs4,caption=The Monitor Superclass]
import java.util.Observable
class Monitor extends Observable

  properties public
  logger_ = Logger.getLogger(Monitor.class.getName())
  sleeptime

  properties static
  da	  = TheDataAccess null
  
  method Monitor()
    this.da  = TheDataAccess.getInstance()
\end{lstlisting}

\begin{lstlisting}[label=obs3,caption=an Observable Thread]
class TransactionStatusMonitor implements Runnable extends Monitor
  
  method TransactionStatusMonitor(s) signals ClassNotFoundException
    this.sleeptime = s
    
  method run()
    do
      loop forever
	pi		  = this.da.idealq1()
	if pi.getID().length() < 5 then
	  do
	    nop
	  end
	else do
	  pi.setIdType('ideal')
	  ibidp		  = this.da.getIBPostIDStatuses(pi)
	  setChanged()
	  notifyObservers(ibidp.getStatusDelta())
	end
	
	Thread.currentThread().sleep(this.sleeptime) -- sleep for sleeptime seconds
      end
    catch InterruptedException
      parse source s
      say "thread interrupted:"  s
    end

\end{lstlisting}
In lines 17 an 18 the magic happens: the \code{setChanged()} method
sets the status of this instance as updated, and the
\code{notifyObservers()} method calls for all the registered
\emph{Observers} their \code{update()} methods; this has the following
signature:

\begin{lstlisting}[label=obs4,caption=the update() method]
  method update(o=Observable,obj=Object) protect
    cl = o.getClass().getName()
\end{lstlisting}
The \code{update()} method receives an \keyword{Object}. The
\code{.getClass.getName} call is for illustrative purposes and can be
used to decide how to treat the received update object.



\chapter{Incorporating Class Libraries}
\section{A Word About Java Generics}
Many classes in Java are expressed as generics. It is important to note that the generic is a \emph{compile time only} java type 
enforcement mechanism, and therefore does not affect \nr{}.

A generic class has, underlying it, a class that accepts one or more objects as parameters - taking as an example the \keyword{ArrayList} class, the Java documentation shows that this has a class signature of \keyword{public class ArrayList<E>} with one of the constructors being \keyword{ArrayList()} and, for example, a method \keyword{add(E e)}.
 If the \keyword{Arraylist} is instantiated in Java as follows:-

\begin{verbatim}
		ArrayList<String> stringList = new ArrayList<String>();
\end{verbatim}

then the Java compiler will note that the ArrayList is instantiated with a \keyword{<String>} object - and will enforce String usage everywhere else that the \keyword{<E>} is used in the class documentation - in this case the type \keyword{add(E e)}.

Thus 
\begin{verbatim}
		stringList.add("Item");
\end{verbatim}
will be permitted by the compiler, since a string is being added. In contrast, 
\begin{verbatim}
		stringList.add(new Integer(7));
\end{verbatim}
will fail since a string is not being added.

Remembering that the \keyword{ArrayList} deals directly with objects the following short \nr{} program will correctly use \keyword{ArrayList} without worrying about the "complication" of generics.

\begin{lstlisting}[label=ArrayListExample, caption=ArrayList Example]
	a1 = ArrayList()	-- An ArrayList just deals with Objects

	a1.add("Eric")		-- so we give it some Rexx objects
	a1.add("Erica")
	num = 0
	a1.add(num)

	say "There are" a1.size "elements in the list" -- and show they are present

	/* Now, to retrieve them */

	loop item over a1
		say item
	end
\end{lstlisting}

If one does not need generics, then it could be asked why they have been implemented at all - the answer is that they prevent many Java run-time errors resulting from a failure to cast the object used to the correct type. When programming in \nr{} the use of the "universal" \keyword{Rexx} class means that this is rarely an issue. When retrieving objects from a generic class used from within Java one must remember to use the correct type, cast or the \keyword{binary} option just as would be expected when using a Java object in any other way.


\section{The Collection Classes}
The Java collections framework (JCF) is a set of classes and interfaces that implement commonly reusable collection data structures.
The JCF provides both interfaces that define various
collections and classes that implement them. Collection
implementations in pre-JDK 1.2 versions of the Java platform included
few data structure classes, but did not contain a collections
framework. The standard methods for grouping Java objects were via the array, the Vector, and the Hashtable classes, which were not easy to extend, and did not implement a standard member interface.
The collections framework was designed and developed primarily by
Joshua Bloch, and was introduced in JDK 1.2.

Almost all collections in Java are derived from the
\keyword{java.util.Collection} interface. Collection defines the basic parts of
all collections. The interface states the \keyword{add()} and \keyword{remove()} methods
for adding to and removing from a collection respectively. Also
required is the \keyword{toArray()} method, which converts the collection into a
simple array of all the elements in the collection. Finally, the
\keyword{contains()} method checks if a specified element is in the
collection. The Collection interface is a subinterface of
\keyword{java.util.Iterable}, so any Collection is iterable (using an
iterator for a loop over the contents).
All collections have an iterator that goes
through all of the elements in the collection.

The Collection framework is one of the aspects of where \nr{}
relegates to Java for its implementation. Where oo\Rexx{} has had its
collection classes in the language definition from day one, in \nr{}
they are not part of the language; most of the data related support is
in the indexed strings feature. This, in turn, makes use of the
Dictionary mechanism already implemented in the earliest versions of Java;
\nr{} language design was long complete when JDK 1.2 came out.

The Pre-Java Generic classes JFC had, in order to be generic, an
interface in which objects could be added in as a java.lang.Object,
but on return, that object needed to be typecast to the right type.
Using collection classes did entail a good deal of casting return
values, as type \Rexx{} was not part of the set of types that
collections had native support for. Modern \nr{} versions
\marginnote{\color{gray}3.02} have builtin support for using type
\Rexx{} in collection classes\footnote{In actuality, the needed
  interfaces, like \emph{Comparable} and \emph{Comparator} are now
  provided in the \Rexx{} type}, so these can be added to and
retrieved from collection classes without further ado. 

The NetRexx native \Rexx{} datatype contains a Java Hashtable which is
part of the Collections Framework. New classes, constructors and
methods have been added to implement the Java Map interface and allow
better interoperation with Java. Some of the new collections support
methods include \keyword{isindexed()} to check if a Map currently exists,
\keyword{size()} to determine the count of map entries and
\keyword{buildmap(sequence1,sequence2)} to construct Rexx maps from arrays or
Java Lists. Other classes and methods are documented in the Java
Collections Map interface Javadocs. "isindexed()" returns 0 if no
indexed values exist and 1 if there is at least one indexed value in a
Rexx object. To build a new indexed Rexx map with the buildmap method
you can do this: \keyword{Rexx(default).buildMap(keys, values)} where
keys and values are any arrays or Java collections framework Lists and
default is the default value for the Rexx variable (using the standard
Rexx constructors). 

All elements are converted to strings before being added to the indexed Rexx variable which is returned. Null can be passed for one of the keys or values parameters to default to a 1-n integer sequence matching the other parameter but if both parameters are provided they must have the same length. Note that arrays do not need to be string arrays and that primitive arrays such as int[] are also accepted.
\begin{shaded}
Collection is a Java generic. Any collection can be written to store any
class. For example, Collection<String> can hold strings, and the
elements from the collection can be used as strings without any
casting required. \nr{} 3.02 added \keyword{loop over} support in
\nr{} programs for collection classes; this has been implemented
without the need for Java generics. This makes it impossible to use
the generics mechanism to constrain collection class membership to a
specified type. This, however, can be easier accomplished by
subclassing the collection class and overriding its constructors.
\end{shaded}

\include{streamio}
\include{javaio}
 

\chapter{Algorithms in \nr}{}
\section{Factorial}
A \emph{factorial} is the product of an integer and all the integers
below it; the mathematical symbol used is ! (the exclamation mark). For
example 4! is equal to 24 (because 4*3*2*1=24). The
following program illustrates a recursive (a method calling itself)
and an iterative approach to calculating factorials.
\lstinputlisting[label=factorial,caption=Factorial]{../../../../examples/rosettacode/RCFactorial.nrx}
Executing this program yields the following result:
\begin{verbatim}
===== Exec: RCFactorial =====
Input a number: 
42
42! = 1405006117752879898543142606244511569936384000000000 (using iteration)
42! = 1405006117752879898543142606244511569936384000000000 (using recursion)
\end{verbatim}
As you can see, fortunately, both approaches come to the same
conclusion about the results. In the above program, both
approaches are a bit intermingled; for more clarity about how to use
recursion, have a look at this:
\begin{lstlisting}[label=factorialrecursive, caption=Factorial Recursive]
class Factorial
numeric digits 64

  method main(args=String[]) static
    say factorial_(42)

  method factorial_(number) static
    if number = 0 then return 1
    else return number * factorial_(number-1)

\end{lstlisting}
In this program we can clearly see that the \texttt{factorial\_} method, that
takes an argument \texttt{number} (which is of type \Rexx{} if we do not
specify it to be another type), calls itself in the method body. This
means that at runtime, another copy of it is run, with as argument
number that the first invocation returns (the result of 42*41), and so
on.

In general, a recursive algorithm is considered more elegant, while an
iterative approach has a better runtime performance. Some language
environments are optimized for recursion, which means that their
processors can spot a recursive algorithm and optimize it by not
making many useless copies of the code. Some day in the near future
the JVM will be such an environment. Also, for some problems, for example
the processing of tree structures, using a recursive algorithm seems
much more natural, while an iterative algorithm seems complicated or forced.
\section{Fibonacci}
\lstinputlisting[label=fibonacci,caption=Fibonacci]{../../../../examples/rosettacode/RCFibonacciSeq.nrx}
\chapter{Using Parse}
The \texttt{Parse} statement is one of the stalwarts of the Rexx
family of languages, and allows one to easily split a string into parts without needing to resort to more traditional techniques of string processing.

The syntax of a parse statement is 

\begin{verbatim}
parse term template 
\end{verbatim}
where \texttt{term} is a string or a previously initialised variable. The template is a list of instructions describing how to split the string.

\section{Literal Parsing}
The most common use of \texttt{parse} is to split a string up into parts separated with a delimiter - whilst the most common delimiter is a simple space any string may be used:-

\begin{lstlisting}[label=SimpleParseExample, caption=Simple Parse Example]
log = "2014/05/15 21:35:47.012 - error in {{[findit]}}"
parse log year "/" month "/" day hour ":" minute ":" second "." msecond "-" text
say "On day" day "of month" month "at about" hour":"minute "we got" text
parse text "{{[" name "]}}"
say name
\end{lstlisting}

Here \texttt{log}  is composed of a datestamp separated from a message by a hyphen. The datestamp is composed of a date separated from a time by a space - within the date the year month and day are delimited by a slash and within the date the hour, minute and second fields by a colon. The millisecond field is separated from the seconds by a decimal point.

The first \texttt{parse} divides these using the relevant delimiter - where there is no delimiter then a space is used.

The \texttt{term} is the variable \texttt{log} and the \texttt{template} is 
\begin{verbatim}
year "/" month "/" day hour ":" minute ":" second "." msecond "-" text
\end{verbatim}

This first template may be read as the following sequence of actions
\begin{enumerate}
\item Assign the contents of \texttt{log} to the variable \texttt{year} until a \texttt{/} is encountered (\texttt{2014})
\item Following the \texttt{/} assign \texttt{month} with the sting found up until another \texttt{/} (\texttt{05})
\item Place the contents following the \texttt{/} until a space into the variable \texttt{day} (\texttt{15})
\item Following the space, assign the value found up until the \texttt{:} into the hour variable (\texttt{21})
\item Repeat for the variable \texttt{minute} (\texttt{35})
\item Assign the \texttt{second} value up until the \texttt{.}
\item Take the value for \texttt{msecond} until a delimiter of \texttt{-} is seen
\item Assign the remainder to variable \texttt{text}
\end{enumerate}

The second \texttt{parse} statement shows how the delimiters can be more complex - the \texttt{template} is 
\begin{verbatim}
"{{[" name "]}}"
\end{verbatim}
and extracts the value between \{\{[ and ]\}\} to the variable \texttt(name)

Running the above example will produce the following output:-
\begin{verbatim}
At about 21:35 we got  error in {{[findit]}} 
findit 
\end{verbatim}

As another example, consider
\begin{lstlisting}[label=WordSplitExample, caption=Parse Word-Split Example]
quote = "Now is the winter of our discontent"
loop forever
	parse quote word quote
	say word
	if quote = "" then leave
end
\end{lstlisting}
This will take the first word from \texttt{quote}, and assign the remainder back into \texttt{quote}, print the word taken and repeat until the variable \texttt{quote} is the empty string. The output from this will be
\begin{verbatim}
Now 
is 
the 
winter 
of 
our 
discontent
\end{verbatim}

\subsection{The Placeholder (dummy) Variable}
The first example assigns values to several variables that are not used - this is unnecessary and can be avoided by the use of a placeholder variable which is the \texttt{.} character.

If this is done, the first parse statement becomes
\begin{verbatim}
parse log . "/" month "/" day hour ":" minute ":" . "." . "-" text
\end{verbatim}
The output will remain the same.

\section{Positional Parsing}

Whilst the majority of parsing can be done using a fixed literal delimiter, the \texttt{parse} instruction also allows parsing based on positional patterns. This is achieved with the use of numerical values in the template - the values may also take a prefix of \texttt{+}, \texttt{-} or \texttt{=}

\begin{description}
\item[\texttt{no prefix} or \texttt{=}] indicates that the number is an \textbf{absolute} column value in the string being parsed
\item[\texttt{+}] indicates a \textbf{relative} position that starts from the specified position \textit{after} the position where the last match occurred
\item[\texttt{-}] indicates a \textbf{relative} position that starts from the specified position \textit{before} the last match
\end{description}

These points are best illustrated by example
\begin{lstlisting}[label=positionalParsing, caption=Positional Parsing]
quote = "Now is the winter of our discontent"
tens  = "         11111111112222222222333333"
units = "12345678901234567890123456789012345"

say quote
say tens
say units

parse quote 10 str1 20 -8 str2 +6 str3
-- str1 starts at column 10 and is 10 chars long
say  str1 "("str1.length")"
-- str2 steps back 8 chars and is 6 chars long
say str2 "("str2.length")"
-- str3 is the remainder of the string (as should be expected)
say str3
\end{lstlisting}
Running this gives the following
\begin{verbatim}
Now is the winter of our discontent 
         11111111112222222222333333 
12345678901234567890123456789012345 
e winter o (10) 
winter (6) 
 of our discontent 
\end{verbatim}

Both \texttt{literal} and \texttt{positional} parsing can be combined. Keen-eyed readers will have noted that the output from the first example contained an extra space before the word \texttt{error}
\begin{verbatim}
At about 21:35 we got  error in {{[findit]}} 
Extra space here     ^^
\end{verbatim}
This is the result of assigning the \textit{remainder} of the string to the variable text - leading blanks are normally removed \textit{except} in this special case.

One can use a positional pattern to eliminate this extra space:-
\begin{lstlisting}[label=combinedParsing, caption=Combined Parsing]
log = "2014/05/15 21:35:47.012 - error in {{[findit]}}"
parse log . "/" month "/" day hour ":" minute ":" . "." . "-" +2 text
say "On day" day "of month" month "at about" hour":"minute "we got" text
parse text "{{[" name "]}}"
say name
\end{lstlisting}

Note that the relative positional pattern used here is \texttt{+2} - \texttt{0} is the position of the last match which is the hyphen, \texttt{+1} is the position of the following space and thus \texttt{+2} is the start of the target string. 

\section{Variable Templates}
Variables may be used as the pattern in the \texttt{templates} in order to accommodate the occasions when the pattern may need to be specified at runtime. An illustration of this is the following evolution of the first example that will correctly parse dates specified in two distinct ways
\begin{lstlisting}[label=variablePattern, caption=Variables in Patterns]
log = ""
log[1] = "2014/05/15 21:35:47.012 - error in {{[findit]}}"
log[2] = "2014-05-15 21:35:47.012 - error in {{[findit]}}"

loop i = 1 to 2
	dtsep = log[i].substr(5,1)
	parse log[i] . (dtsep) month (dtsep) day hour ":" minute ":" . "." . "-" +2 text
	say "On day" day "of month" month "at about" hour":"minute "we got" text
end
\end{lstlisting}
Note that he date separator \texttt{dtsep} is determined and then used in the parse pattern by enclosing it in parentheses, thus \texttt{(dtsep)}. The output of this program is
\begin{verbatim}
On day 15 of month 05 at about 21:35 we got error in {{[findit]}} 
On day 15 of month 05 at about 21:35 we got error in {{[findit]}} 
\end{verbatim}
It can be seen that the date was successfully parsed in both cases.

It is important to note that any pattern specified by a variable \textit{will be assumed to be literal unless it has a +, - or = prefix}. Should one wish to use positional patterns then the prefix \textbf{must} be used.
\begin{lstlisting}[label=variablePosPattern, caption=Variables in Positional Patterns]
message = "this is a message that contains the number 10- just there, see?"
pat = "10"
parse message part1 5 (pat) part2
say "literal:" part1 part2
parse message part1 5 =(pat) part2
say "positional:" part1 part2
\end{lstlisting}
When run this illustrates the difference between the two parse statements
\begin{verbatim}
literal: this - just there, see? 
positional: this  message that contains the number 10- just there, see?
\end{verbatim}
\chapter{Using Trace}
The \texttt{trace} command is the inbuilt debugging facility of the \Rexx{} family, and, as might be expected from its name, allows one to trace the execution of your program. It is possible to trace both program statements and the state of variables within your code.

\texttt{(Trace)} is a compile-time option, and should be disabled once debugging as been completed.

The syntax of the trace command is
\begin{verbatim}
trace traceitem 
\end{verbatim}
where \texttt{traceitem} defines the behaviour of the trace command. Only one \texttt{traceitem} may be given, and only one of the program statement tracing options will be in use at any time. 
Variable tracing options, however, are \textit{additive} and such statements may appear multiple times.

All \texttt{trace} output is headed by three hyphens followed by the source file name, as follows
\begin{verbatim}
--- TerribleExample.nrx
\end{verbatim}

\section{Tracing Program Statements}
The \texttt{traceoptions} that affect the tracing of program statements are
\begin{description}
\item[\textbf{all}] will display all statements as they are executed. Each line in the trace output will be prefixed with \texttt{*=*} or a \texttt{*-*} should output span subsequent lines.

The \texttt{trace all} statement can be placed anywhere in the program source. 

\item[\textbf{methods}] will show the each method as it is invoked, along with any parameters to it. The trace output for method traces is prefixed by a \texttt{*=*} for the method call itself and a \texttt{>a>} indicating the assignment of a value to a method parameter. \textit{No other program statements will be traced.}

The \texttt{trace methods} statement should be placed \textit{before} the first method is defined in a class.

\item[\textbf{results}] acts as though the \texttt{trace all} statement had been given, and, if placed \textit{before} any method will also act as though \texttt{trace methods} was also specified.

In addition to the \texttt{all} and \texttt{methods} tracing implied by \texttt{results} the following will also take place
\begin{description}

\item[Properties] will have their assignments shown. These will be identified by \texttt{>p>}

\item[Local variables] will also be traced, with assignments identified by \texttt{>v>}

\item[Expressions] will have their evaluations shown if not shown for as a part of \texttt{properties} or \texttt{local variable} trace output. Such evaluations are indicated by \texttt{>>>}
\end{description}
\item[\textbf{int}] will stop execution of the \nr{} program, allow single stepping through the program and provide read and write access to variables and properties. 
To leave the interactive trace type \texttt{trace off} at the \texttt{*->} prompt.
\item[\textbf{off}] \texttt{trace off} disables tracing. No further tracing output will take place.
\end{description} 

\section{Tracing Variables}
The all-or-nothing tracing offered by, for instance \texttt{trace results} can lead to a deluge of trace information in many cases.

In these instances one may more finely control which variables one wishes to monitor using the \texttt{trace var} statement.
The syntax of the \texttt{trace var} statement is 
\begin{verbatim}
trace var var1 [var2...]
\end{verbatim}
or
\begin{verbatim}
trace var -var1 [-var2...]
\end{verbatim}
where the first form adds variables to the list that should be watched, and the second removes them. The forms may be mixed to add some variables and remove others simultaneously, as here:-
\begin{verbatim}
trace var var1 -var2 var3 -var4 -var5
\end{verbatim}
to  monitor \texttt{var1} and \texttt{var3} and remove \texttt{var2, var5} and \texttt{var5} from the list of watched variables. 

Multiple \texttt{trace var} statements may be used, as mentioned above.

It is not an error to specify a variable name that does not exist.

Each variable can appear only \textit{once} in a \texttt{trace} statement.

A variable name may that of any type - including arrays (without the \texttt{[]}).

Program tracing options never alter the list of watched variables. If tracing has previously been turned off then variable tracing may be resumed simply with a \texttt{trace var} statement.

\section{Interactive tracing}
\index{Interactive tracing}

Interactive tracing is activated by the \keyword{trace int} instruction, 
which will cause the interpreter to stop and present the interactive 
trace prompt. 

Pressing [Enter] continuously, single-steps and executes 
every clause in the program. Variables and properties are accessible in 
both read and write mode.

Typing '?' shows the interactive trace capabilities:
\begin{lstlisting}
    *-> ?
Experimental interactive trace :
  press [Enter] to trace interactively
  type '=' to reinterpret current clause
  type '-[n]' to show previous n clause(s), shows current clause if n is absent
  type '+[n]' to show next n clause(s), shows next clause if n is absent
  type 'trace off' to stop tracing
  any other clause entered must be either an assignment or a SAY instruction
    *-> 
\end{lstlisting}

Given below trace.nrx program, 
\begin{lstlisting}
class trace
  properties public static
  p = 1
 
  method main(args=String[]) static
    say 'interactively tracing..'
    trace int
    i=0
    say 'i='i 
    i=traced(i)
    say 'Hello world'||'!'.copies(i)

  method traced(arg=Rexx) static
    say 'traced called with arg:'arg  
    return arg+p
\end{lstlisting}

a sample interactive trace session might look as follows:


\begin{lstlisting}
$ nrc -exec trace
NetRexx portable processor 4.04
Copyright (c) RexxLA, 2011,2022.   All rights reserved.
Parts Copyright (c) IBM Corporation, 1995,2008.
Program trace.nrx
  === class trace ===
    function main(String[])
    function traced(Rexx)
===== Exec: trace =====
interactively tracing..
    *->                                       -- interrupted at trace int, single-step
    --- trace.nrx
  8 *=*     i=0
    >v> i "0"
    *->                                       -- single-step
  9 *=*     say 'i='i 
    >>> "i=0"
i=0
    *-> i=2                                   -- update variable i
    *-> =                                     -- re-execute current clause
    >>> "i=2"
i=2
    *-> say 'p='p                             -- say something
p=1
    *-> p=2                                   -- update property p
    *-> say 'p='p
p=2
    *->                                       -- single-step
 10 *=*     i=traced(i)                       
    >>> "2"
traced called with arg:2
    >v> i "4"
    *->                                       -- single-step
 11 *=*     say 'Hello world'||'!'.copies(i)
    >>> "4"
    >>> "Hello world!!!!"
Hello world!!!!
    *-> i=-1                                  -- update variable
    *-> i=traced(i)                           -- call method
traced called with arg:-1
    *-> say 'i='i                             -- say something
i=1
    *-> =                                     -- re-execute
    >>> "1"
    >>> "Hello world!"
Hello world!
    *-> trace off                             -- trace off
Processing of 'trace.nrx' complete

\end{lstlisting}

\section{Examples}
\subsection{Program Trace}
\subsubsection{Trace All}
Running the program below
\begin{lstlisting}[label=Trace1, caption=Trace Example 1]
trace all

class traceExample   

	properties
		aIs
		bIs
	
	method traceExample(a, b)
		aIs = a
		bIs = b
		
	method times
		return aIs * bIs

	method main($cmdin1=String[]) static
		arg=Rexx($cmdin1)
		te = traceExample(2, 3)
		fred = te.times
		say fred
\end{lstlisting}
gives trace output of
\begin{verbatim}
     --- traceExample.nrx 
  16 *=*  method main($cmdin1=String[]) static
     >a> $cmdin1 "[Ljava.lang.String;@72ebbf5c" 
  17 *=*   arg=Rexx($cmdin1)
  18 *=*   te = traceExample(2, 3)
   9 *=*  method traceExample(a, b)
     >a> a "2" 
     >a> b "3" 
  10 *=*   aIs = a
  11 *=*   bIs = b
  12 *-* 
  19 *=*   fred = te.times
  13 *=*  method times
  14 *=*   return aIs * bIs
  20 *=*   say fred
\end{verbatim}
This output may be read thus
\begin{description}
\item[\hphantom{111}--- traceExample.nrx] Identification of the program being traced. This is the \texttt{tracing context}.
\item[16 *=*  method main(\$cmdin1=String[]) static] The first line that is actually executed is line 16.
\item[\hphantom{111}>a> \$cmdin1 "[Ljava.lang.String;@72ebbf5c"] Variable \texttt{\$cmdin1} is assigned a string value from the java virtual machine.
\item[17 *=*   arg=Rexx(\$cmdin1)]Line 17 is executed next...
\item[18 *=*   te = traceExample(2, 3)] followed by line 18
\item[\hphantom{1}9 *=*  method traceExample(a, b)] Line 18 is a method call to a method on line 9...
\item[\hphantom{111}>a> a "2"] which assigns a value of 2 to parameter a
\item[\hphantom{111}>a> b "3"] and a value of 3 to parameter b
\item[10 *=*   aIs = a] the following lines document only code execution
\item[11 *=*   bIs = b] 
\item[12 *-*] 
\item[19 *=*   fred = te.times]
\item[13 *=*  method times]
\item[14 *=*   return aIs * bIs]
\item[20 *=*   say fred]
\end{description}

\subsubsection{Trace Methods}
Replacing the \texttt{trace all} from line 1 with \texttt{trace methods} gives trace output of
\begin{verbatim}
    --- traceExample.nrx 
 16 *=*  method main($cmdin1=String[]) static
    >a> $cmdin1 "[Ljava.lang.String;@8094cc7" 
  9 *=*  method traceExample(a, b)
    >a> a "2" 
    >a> b "3" 
 13 *=*  method times
\end{verbatim}
As should be expected, this is a subset of the output provided when using \texttt{trace all}.
\subsubsection{Trace Results}
Replacing the \texttt{trace all} from line 1 with \texttt{trace results} would give
\begin{verbatim}
    --- traceExample.nrx 
 16 *=*  method main($cmdin1=String[]) static
    >a> $cmdin1 "[Ljava.lang.String;@72ebbf5c" 
 17 *=*   arg=Rexx($cmdin1)
    >>> "[Ljava.lang.String;@72ebbf5c" 
    >v> arg "" 
 18 *=*   te = traceExample(2, 3)
    >>> "2" 
    >>> "3" 
  9 *=*  method traceExample(a, b)
    >a> a "2" 
    >a> b "3" 
 10 *=*   aIs = a
    >p> aIs "2" 
 11 *=*   bIs = b
 12 *-* 
 11 >p> bIs "3"
 18 >v> te "traceExample@53606bf5"
 19 *=*   fred = te.times
 13 *=*  method times
 14 *=*   return aIs * bIs
    >>> "6" 
 19 >v> fred "6"
 20 *=*   say fred
    >>> "6" 
\end{verbatim}
Here is can be seen that more information is available. Noticeably, the values of assignments are given.
For instance
\begin{description}
\item Line 17 now has an entry of \textbf{>v> arg ""} showing that hte value of the variable \texttt{arg} was set to the empty string
\item Line 18 now has the values of the specified parameters evaluated (\mbox{\textbf{\textgreater\textgreater> "2"}} and \mbox{\textbf{\textgreater\textgreater>> "3"}})
\item Lines 10 and 11 show that values were assigned to parameters (\mbox{\textbf{>p> aIs "2"}} and \mbox{\textbf{>p> bIs "3"})}
\item Line 18 then shows the assignment of the instantiated class to variable \texttt{te}
\item Line 14 shows the evaluation of the multiplication (\mbox{\textbf{\textgreater\textgreater> "6"}}), which is assigned to variable \texttt{fred} in line 19 (\mbox{\textbf{>v> fred "6"}}) on line 19.
\item Finally we see the evaluation of variable \texttt{fred} on line 20.
\end{description}
\subsection{Variable Tracing}
Consider the following example:-
\begin{lstlisting}[label=Trace2, caption=Trace Example 2]
a = "a"
b = "b"
c = 1
d = 2
e = 3

trace var a b c d e f y
z = a || b
y = c + d
f = y + 2
e = f

trace var -a -c -d -e 
y = y * 2
a = y
e = a
\end{lstlisting}
Running this will produce the output below
\begin{verbatim}
    --- variableTraceExample.nrx 
  9 *=* y = c + d
    >v> y "3" 
 10 *=* f = y + 2
    >v> f "5" 
 11 *=* e = f
    >v> e "5" 
 14 *=* y = y * 2
    >v> y "6" 
\end{verbatim}
It can be seen that only the lines that contain watched variables are traced. This the variable assignments on lines 9, 10 and 11 are displayed, since the variables being watched from line 7 to line 12 are \texttt{a, b, c, d, e, f} and \texttt{y}.

Following this, however only the assignment to variable \texttt{y} is shown, since the variables \texttt{a, b ,c d} and \texttt{e} are removed from the list with the command \mbox{\texttt{trace var -a -c -d -e}}.


\section{Tracing Notes}
One further prefix may be encountered in the trace outout \texttt{+++} which signifies an error.

Whenever tracing transfers to a different source file, a new \texttt{tracing context}, identified by the \texttt{---} prefix is output.

Tracing is expensive, and may dramatically impact the run-time performance of the program being traced. Judicious use may therefore be warranted.

\chapter{Concurrency}
\section{Threads}
Threads are a built-in multitasking feature of the JVM. Where earlier
JVM implementations sometime ran on so-called \emph{Green Threads},
which is a library that implements thread support for OS'ses that do
not have this facility (an
early version of Java was called \emph{GreenTalk} for this reason), modern versions
all use native OS thread support. 

A new thread is created when we create an instance of the Thread class. We cannot tell a thread which method to run, because threads are not references to methods. Instead we use the Runnable interface to create an object that contains the run method:

Every thread begins its concurrent life by executing the run method. The run method does not have any parameters, does not return a value, and is not allowed to signal any exceptions.
Any class that implements the Runnable interface can serve as a target of a new thread. An object of a class that implements the Runnable interface is used as a parameter for the thread constructor.

Threads can be given a name that is visible when listing the threads in your system. It is good practice to name every thread, because if something goes wrong you can see which threads are still running.
Additionally, threads are grouped by thread groups. If you do not
supply a thread group, the new thread is added to the thread group of
the currently executing thread. The threads of a group and their
subgroups can be destroyed, stopped, resumed, or suspended by using
the ThreadGroup object. 

The next two samples are used in the following programs that
illustrate thread usage.
\lstinputlisting[label=threads1,caption=Thread sample 1]{../../../../examples/NrxRedBk/thread/ThrdTst1.nrx} 
\lstinputlisting[label=threads2,caption=Thread sample
2]{../../../../examples/NrxRedBk/thread/ThrdTst2.nrx} 
The second class, Hello2, does not \emph{implement} the
\texttt{Runnable} interface, but subclasses it, so it inherits its
methods. This is a valid approach, and it is up to the developer to
choose an implementation and worry about the semantics of an inherited
thread interface.
A newly created thread remains idle until the start method is invoked. The thread then wakes up and executes the run method of its target object. The start method can be called only once. The thread continues running until the run method completes or the stop method of the thread is called.

\include{nrgochannel}

\include{nruappl}
% \chapter{Network Programming}
% \section{Using Uniform Resource Locators (URL)}
% \section{TCP/IP Socket I/O}
% \section{RMI: Remote Method Interface}

\chapter{Database Connectivity with JDBC}
For interfacing with Relational Database Management Systems (RDBMS)
\nr{} uses the Java Data Base Connectivity (JDBC) model. This means
that all important database systems, for which a JDBC driver has
been made available, can be used from your \nr{} program. This is a
large bonus when we compare this to the other open source scripting
languages, that have been made go by with specific, nonstandard
solutions and special drivers. In contrast, \nr{} programs can be
made compatible with most database systems that use standard SQL, and,
with some planning and care, can switch database implementations at
will.
\lstinputlisting[label=jdbc,caption=A JDBC Query example]{../../../../examples/NrxRedBk/jdbc/JdbcQry.nrx} 

The first peculiarity of JDBC is the way the driver class is
loaded. When most classes are 'pulled in' by the translator, a JDBC
driver traditionally is loaded through the reflection API. This
happens in line 22 with the \texttt{Class.forName} call. This implies
that the library containing this class must be on the classpath.

\begin{shaded}\noindent
In previous versions of JDBC, to obtain a connection, one first had to
initialize the JDBC driver by calling the method Class.forName. Any
JDBC 4.0 drivers that are found on the class path are automatically
loaded. (However, one must manually load any drivers prior to JDBC 4.0
with the method Class.forName.)
\end{shaded}\indent

In line 32 of the example we connect to the database using a url and a
userid/password combination. This is an easy way to do and test, but
for most serious applications we do not want plaintext userids and
passwords in the sourcecode, so most of the time we would store the
connection info in a file that we store in encrypted form, or we use
facilities of J2EE containers that can provide data sources that take
care of this, while at the same time decoupling your application
source from the infrastructure that it will run on.

In line 47 the query is composed by filling in variables in a Rexx
string and making a \texttt{Statement} out of it, in line 50. In line
  55, the \texttt{Statement} is executed, which yields a
  \texttt{ResultSet}. This has a \emph{cursor} that moves forward with
  each \texttt{next} call. The \texttt{next} call returns \emph{true}
  as longs as there are rows from the resultset to return.

The \texttt{ResultSet} interface implements \emph{getter} methods for
  all JDBC Types. In the above example, all returned results are of
  type \texttt{String}.

\lstinputlisting[label=jdbcu,caption=A JDBC Update
example]{../../../../examples/NrxRedBk/jdbc/JdbcUpd.nrx} 
For database updates, we connect using the driver in the same way
(line 23) and now prepare the statement used for the database update
(line 50). In this example, we loop through the cursor of a select
statement and update the row in line 66. The \texttt{executeUpdate}
method of \texttt{PreparedStatement} returns the number of updated
rows as an indication of success.

From JDBC 2.0 on, cursors are updateable (and scrollable, so they can
move back and forth), so we would not have to go
through this effort - but it is a valid example of an update statement.
\chapter{WebSphere MQ}
WebSphere MQ (also and maybe better known as MQ Series) is IBM's
messaging and queing middleware, and is in use at a great many financial
institutions and other companies. It has, from a programming point of
view, two API's: JMS (Java Messaging Services), a generic messaging
API for the Java world, and MQI, which is older and proprietary to
IBM's product. The below examples show the MQI; other examples might
show JMS applications.

This is the sample Java application for MQI, translated (and a lot
shorter) to NetRexx.
\lstinputlisting[label=mqsample,caption=MQ Sample]{../../../../examples/enterprise/wmq/MQSample.nrx}
This sample connects to the Queue Manager (called \emph{rjtestqm}) in
\emph{bindings mode}, as opposed to \emph{client mode}. Bindings mode
is only a connection possibility for client programs that are running
in the same OS image as the Queue Manager, on the server. Note that
the application connects (line 19), accesses a queue (line 23), puts a
message (line 32), gets it back (line 39) closes the queue (line 45) and disconnects (line 48) all without checking
returncodes: the exceptionhandler takes care of this, and all
irregulaties will be reported from the catch MQException block
starting at line 50).

The main method does in this case not follow the canonical form, but
has 'binary' as an extra option. Option binary can be defined on the
command line as an option to the translator, as a program option, as a
class option and as a method option. Here the smallest scope is
chosen. There is a good reason to make this method a binary method:
accessing a queue in MQ Series requires some options that are set
using a mask of binary flags - this works, in current \nr{} versions,
only in binary mode, because the operators have other semantics in
nobinary mode.

\lstinputlisting[label=mqlistener,caption=MQ Message Reader]{../../../../examples/enterprise/wmq/MessageReader.nrx}
In contrast to the previous sample the MessageReader sample only has
one import statement. This is always hotly debated in project teams,
one school likes the succinctness of including only the top level
import, and only goes deeper when there is ambiguity detected; another
school spells out the all imports to the bitter end. 

The MessageReader sample connects to another queue, called TESTQUEUE1
(specified in line 7) but here we connect in \emph{client mode}, as
indicated by lines 13-15 which specify an MQEnvironment. Other
options are using an MQSERVER environment variable or a \emph{Channel
Definition Table}.

This program is also uncommon in that it uses
\texttt{MQConstants.MQGMO\_WAIT} as an option instead of being
triggered as a process by a message on a trigger queue. Using this
option means that the program waits (stays active, not really busy polling
but depending on an OS event) until a new
message arrives, which will be processed immediately.

In lines 18-21 a \emph{Channel Exit} is specified. This exit is show
in the following example.
\lstinputlisting[label=mqjavachannelexit,caption=MQ Java Channel Exit]{../../../../examples/enterprise/wmq/TimeoutChannelExit.nrx}
 \lstinputlisting[label=watchdogtimer,caption=WatchdogTimer]{../../../../examples/enterprise/wmq/WatchdogTimer.nrx}
MQ Series has traditional channel exits (programs that can look at the
message contents before the application gets to it). In the MQI Java
environment there is something akin to this functionality, but a Java
channel exit for MQ Series has to be defined in the application, as
shown in the previous example. The function of this particular exit is
to implement a \emph{Watchdog timer} - on a separate thread, as shown
in the sample that follows the sample channel exit. The timer
threatens here to have issues a HP OpenView alert, but that part has
been left out.

This particular sample has been designed to do something that is
normally a bit harder
to do: signal the operations department when something does NOT happen
- here the assumption is that there is a payment going over the queue
at least once every 20 minutes - when that does not happen, an alert
is issued. With every message that goes through, the timer thread is
reset, and only when it is allowed to time out, action is undertaken.

\lstinputlisting[label=watchdogtimer,caption=Publish/Subscribe]{../../../../examples/enterprise/wmq/MQPubSubSample.nrx}
This sample shows the publish-subscribe interfaces that at some time have been
added to the product. This specific sample shows some Java thread
complexity but is a good example of doing publish/subscribe work in a
multithreaded way, which is a natural fit for this type of work.

\include{mqtt}

\chapter{Component Based Programming: Beans}
JavaBeans is the name for the Java component model. It consists of two
conventions, for the naming of \emph{getter} and \emph{setter} methods
for properties, and the \emph{event} mechanism for sending and
receiving events. \nr{} adds support for the automatic generation of
getter and setter methods, throught the \textbf{properties indirect}
option on the properties statement.

\include{jsr223}

\chapter{\nr{} Tools}
\section{Editor support}\label{editors}
This chapter lists editors that have plugin support for \nr{},
ranging from syntax coloring to full IDE support (specified), and
\Rexx{} friendly editors, that are extensible using \Rexx{} as a macro
language (which can be the first step to provide \nr{} editing support).
\subsection{JVM - All Platforms}
\begin{tabularx}{\textwidth}{>{\bfseries}lX}
\toprule
JEdit & Full support for \nr{} source code editing, to be found at
\url{http://www.jedit.org}.
\\\midrule
NetRexxDE & A revisions with additions of the \nr{} plugin for
jEdit, moving to a full IDE for \nr{}. \url{http://kenai.com/projects/netrexx-misc} 
\\\midrule
Eclipse & Eclipse has a \nr{} plugin that provides a complete IDE
environment for the development of \nr{} programs (in alpha release)
by Bill Fenlason. The project is situated at SourceForge
(\url{http://eclipsenetrexx.sourceforge.net/}).
% Chapter
% \ref{setupeclipse} on page \pageref{setupeclipse} discusses the setup
% of Eclipse to build the translator itself; and has
% instructions for the setup of the \nr{} plugin.
\\\bottomrule
\end{tabularx}
\subsection{Linux}
\begin{tabularx}{\textwidth}{>{\bfseries}lX}
\toprule
Emacs & netrexx-mode.el (in the \nr{} package in the \texttt{tools}
directory) runs on GNU Emacs, which is installed by default on most
Linux developer distributions.
\\\midrule
vim & vi with extensions
\\\bottomrule
\end{tabularx}
\subsection{MS Windows}
\begin{tabularx}{\textwidth}{>{\bfseries}lX}
\toprule
Emacs & netrexx-mode.el (in the \nr{} package in the \texttt{tools} directory) runs on GNU Emacs for
Windows. \url{http://www.gnu.org/software/emacs/windows/faq.html}.
\\\midrule
vim & vi with extensions
\\\bottomrule
\end{tabularx}
\subsection{macOS}
\begin{tabularx}{\textwidth}{>{\bfseries}lX}
\toprule
Aquamacs & A version of Emacs that is integrated with the macOS Aqua
look and feel. (\url{http://www.aquamacs.org}). \nr{} mode is
included in the \nr{} package in the \texttt{tools} directory.
\\\midrule
Emacs & netrexx-mode.el (in the \nr{} package) runs on GNU Emacs for
macOS. \url{http://www.gnu.org/software/emacs}.
\\\midrule
Vim & Vi with extensions
\\\bottomrule
\end{tabularx}
\section{Java to Nrx (java2nrx)}
When working on a piece of Java code, or an example written in the
language, sometimes it would be good if we could see the source in
\nr{} to make it more readable. This is exactly what \emph{java2nrx}
by Marc Remes does. It has a  Java 1.5 parser and an Abstract Syntax
Tree that delivers a translation to NetRexx, to the
extend of what is currently supported under NetRexx.

At the moment it is to be found at \url{git clone git://git.code.sf.net/p/netrexx/code netrexx-code} in the tools directory.

It is started by the \texttt{java2nrx.sh} script; for convenience, place \texttt{java2nrx.sh} and \texttt{java2nrx.jar} in the
same directory. NetRexxC and java must be available on the path.

%% Usage:
%% \input{../../../railroad/java2nrx1}
%% Alternatively:
%% \input{../../../railroad/java2nrx2}
\include{eclipseguide}
\chapter{Platform dependent issues}
\section{Mobile Platforms}
Android\texttrademark is a version of Linux with a runtime consisting
of a variant of Java, and is friendly to \nr{}
programs. Indeed, with \nr{} performing so much better than the closest
competition (jRuby, jython) on these devices, there might be a bright
future for \nr{} in these environments. 

However, there are some drawbacks, caused by the security architecture
put in place. Free, unfettered programming like one can do on a
desktop machine is a rare occurrence on these devices, and to get
programs running on them requires some knowledge of the security
architecture that has been put in place for mobile operating systems.

While Apple development still employs a closed model that allows programming only by
buying a license with accompanying certificates, and vetting by the
App Store employees, and an assumption you will program in
Objective-C, Android allows programming but not as straightforward as
we know it. To make simple command-line \nr{} programs, both device
types need to be \emph{rooted} to allow optimal access. Android allows
the installation of applications without vetting by third parties, but
dictates a programming model that incurs some overhead - which is a
drawback for the occasional scripter.
\subsection{Android}
The security model of Android is based on \emph{least needed
  privilege} and is implemented by assigning each application a
different userid, so that applications on the same device (be it a
phone or a tablet) cannot get to each others data. The consequence of
this is that simple \nr{} programming and scripting on the device itself is limited, however developing complete applications in NetRexx is not.

\subsection{Apple IOS}
There is a number of ways \nr{} can be run on Apple IOS devices
(iPhone and iPad). Both have drawbacks. With ISH, a 32-bit version of
Linux is started on an emulated X86 processor; this has dire
consequences for performance. The 'Jailbreak' solution runs with much
better performance, but this approach is rather volatile and cannot be
guaranteed to be feasible in the future, because Apple is actively
fixing the holes that allow it.
\subsubsection{ISH}
The ISH application delivers a Linux shell on emulated hardware. The
NetRexxC.jar can be transferred with scp to the storage of ISH, from
where it can be run. It needs higher memory heap allocations than the
standard; -Xms128M -Xmx128M is recommended here. Do not expect
performance corresponding to the native ARM hardware in your device.
\subsubsection{Jailbreak}
Note: this chapter is out of date. 
Nonewithstanding the current intention of Apple to only allow Swift an
d Objective-C
as programming languages on the iPhone and iPad, \nr{} on IOS works fine. This is what one should do to make it work:
\begin{enumerate}
\item Jailbreak\footnote{Note that jailbreaking an iPhone is against Apple's End Use License Agreement) and might be illegal in some jurisdictions.} the device. This is necessary until a more sensible setup is used. I used Spirit; it synchs the phone with the hack and then Cydia is installed, an application that does package management the Debian way
\item Choose the "developer profile" on Cydia when asked. This applies a filter to the packages shown (or rather it doesn't) - but you need to do it in order to see the prerequisites
\item OpenTerminal will help you to do command line operations on the phone itself
\item The prerequisites are a Java VM (JamVM installs a VM and
  ClassPath, the open Java implementation) and Jikes, the Java
  compiler written in C and compiled to the native instruction set of
  the phone, which is ARM - most processors implementing this have
  \emph{Jazelle}, a specials instructionset to accelerate Java
  bytecode. However, this feature is seldom used. 
\end{enumerate}
The phone can also be logged on to using ssh from your desktop. Do not forget to change the password for the 'root' user and the 'mobile' user, as instructed in the Cydia package. Note that this type of information will can be made inaccurate very swiftly.

When this is done, NetRexxC.jar can be copied to the phone. I did this using 'scp NetRexxC.jar mobile@10.0.0.76:' (use the password you just set for this userid) (and because my router assigned 10.0.0.76 to the phone today). I crafted a small 'nrc' script that does a translate and then a Java compile using jikes (and I actually wrote this on the phone using an application called 'iEdit' - nano, vim and other editors are also available but I found the keyboard scheme to type in ctrl-characters a bit tedious - you type a 'ball' character and then the desired ctrl char, while shifting the virtual keyboard through different modes):

nrc:
\begin{verbatim}
java -cp ~/NetRexxC.jar COM.ibm.netrexx.process.NetRexxC $*
\end{verbatim}
Now we can do a compile of the customary hello.nrx with './nrc -keep -nocompile hello' (notice that this is all in the home directory of the 'mobile' user, just like the jar that I just copied. The resulting hello.java.keep can then be mv'ed to hello.java and compiled with 'jikes hello.java'. This produces a class that can be run with 'java -cp NetRexxC.jar hello'
\section{IBM Mainframe: Using \nr{} programs in z/OS batch}
Traditionally the mainframe was a batch oriented environment, and much
of the workload that counts still executes in this way. To be able to
use \nr{} with Job Control Language (JCL) in batch address spaces,
accessing traditional datasets and interacting with the console when
needed, we need to know a bit more. This will be explained in these paragraphs.

A standard component of z/OS since version 1.8 or so is \texttt{jzos},
which acts as glue between the unix-like abstractions the JVM works
with and the time tested way of working on z/OS, with its SAM and VSAM
datasets, its Partitioned Data Set (PDS) file organization, the ICF
Catalogs and console address space; all of which in existence long
before Java reared its head in our IT environments.

The manuals will teach you that there are several ways to
interact with HFS/OMVS resources in JCL, but the alternatives to
\texttt{jzos} have so many drawbacks that it is the only
sensible way to run \nr{} programs in the batch environment. 

\subsection{Example}
\begin{lstlisting}
  //AB2217N1 JOB (7355,710,TC78JAN),'PGM',MSGCLASS=X,NOTIFY=AB2217  
//JAVA EXEC PROC=JVMPRC60,                                        
// JAVACLS='HelloWorld'                                           
//STDENV DD *                                                     
. /etc/profile                                                    
export JAVA_HOME=/usr/lpp/java/J6.0                               
export PATH=/bin:"${JAVA_HOME}"/bin                               
LIBPATH=/lib:/usr/lib:"${JAVA_HOME}"/bin                          
LIBPATH="$LIBPATH":"${JAVA_HOME}"/lib/s390                        
LIBPATH="$LIBPATH":"${JAVA_HOME}"/lib/s390/j9vm                   
LIBPATH="$LIBPATH":"${JAVA_HOME}"/bin/classic                     
export LIBPATH="$LIBPATH":                                        
APP_HOME=$JAVA_HOME                                               
CLASSPATH=$APP_HOME:"${JAVA_HOME}"/lib:"${JAVA_HOME}"/lib/ext     
for i in "${APP_HOME}"/*.jar; do                                  
    CLASSPATH="$CLASSPATH":"$i"                                   
    done
export CLASSPATH="$CLASSPATH":                                    
IJO="-Xms16m -Xmx128m"                                            
export IBM_JAVA_OPTIONS="$IJO "                                   
//
\end{lstlisting}
\include{building}
\include{datetimearith}
\include{nrws}

\include{inner}

\include{resolution}

\backmatter
% \listoffigures
% \listoftables
% \lstlistoflistings
\printindex
\clearpage
\psset{unit=1in}
\begin{pspicture}(3.5,1in)
  \psbarcode{\isbn}{includetext guardwhitespace}{isbn}
\end{pspicture}
\end{document} 
