\input{../boilerplate/preamble}
\begin{document}    
\renewcommand{\isbn}{978-90-819090-2-0}    
\setcounter{tocdepth}{1} 
\title{Colofon}
\author{Rene Jansen}
\date{Version 1 of \today}
\maketitle
\pagenumbering{Roman}
\pagestyle{plain}
\frontmatter
\pagenumbering{Roman}
\pagestyle{plain}
\input{../boilerplate/bookmeta}
\tableofcontents
\newpage
\pagenumbering{arabic}
\frontmatter
\large
\input{../boilerplate/series.tex}
\input{../boilerplate/conventions.tex}
\include {introduction}
\mainmatter
\chapter{Structure}
\section{Prerequisites}
There arevery few prerequisites for running the book production
toolchain, at least in comparison to other toolchains. All executables
but one are readily available in compiled form on the net. A few
remarks.
\begin{enumerate}
\item The toolchain uses the \textbf{xelatex} version of Latex, because of
its easy integration of platform fonts and its facility to directly
produce PDF files. This program is available in the \emph{livetex}
distribution, to be found at \url{http://}.
\item The make utility (in the GNU version) must be available for the
  platform. All modern platforms have it included, there is a binary
  for Windows downloadable from \url{http://}.
\item For the utility that produces railroad diagrams, there are
  platform versions checked into the \nr source repository
\item to view the generated
\end{enumerate}

\section{The Makefile}
The directory ../netrexx/netrexxc/trunk/documentation/ contains a lot
of files. There is one subdirectory that contains the main files for
this colofon, it is name \textbf{colofon}. So in
../netrexx/netrexxc/trunk/documentation/colofon there are files we are
concentrating on here. The file to look into to find your starting
point is the file \textbf{makefile}. In the makefile the rules for the
make utility are found. Executing the command
\begin{verbatim}
make -B
\end{verbatim}
causes the book to be rebuilt from its sources. Sources are defined as
files ending in .tex, but there are also prerequisites in the form of
files that are pregenerated from *.tex files in previous runs. The
standard makefile re-runs the commands until it has determined that
all prerequisites have been met - this is at least its intention.

The main sourcefile for this book is called colofon.tex . It imports
different files from a ../boilerplate directory, which contains
materials shared by all books in the series. This is seen in the
import and include tags. It includes the file
\textbf{introduction.tex} to show how files are included, but the text
for these chapters the input is in the files colofon.txt itself. How
and why to split input text files is up to the author. The User's
Guide (Quick Start Guide) for example has a lot if imports/includes,
because the original material contained a lot of separate files. The
Programmer's Guide has less files, but uses the \emph{listings}
package a lot to import \nr source files from the examples section. 
\backmatter
\listoffigures
\listoftables
\lstlistoflistings
\printindex
\clearpage
\psset{unit=1in}
\begin{pspicture}(3.5,1in)
  \psbarcode{\isbn}{includetext guardwhitespace}{isbn}
\end{pspicture}
\end{document} 
